Circumnuclear Media and Accretion Rates of Quiescent Supermassive Black Holes

### 2015-04-27 ms.tex
cutting fat - and lots ofit16

\citet{BergerZauderer+:2012a} infer a flattening of the gas density
profile of the host of J1644+57 on radial scales $r \gtrsim 0.4$ pc
(their Fig.~6) which looks qualitatively similar to the shape density
profile we predict for core galaxies (Fig.~\ref{fig:profiles}, solid
line).  To obtain such a flattening on the inferred radial scales
would require a black hole of $\Mbh\sim 10^7 \Msun$.  However, this is
inconsistent with the break in the gas density seen at $\sim 0.6$ pc
by \citet{BergerZauderer+:2012a}, which should correspond to the break
radius of the {\it stellar density} profile, thus implying an
unrealistically small SMBH mass of $\Mbh\simeq 100\Msun$ according to
the mean $r_b-M_{\bullet}$ relationship \citep{LauerFaber+:2007a}.
Alternatively, the observed density break in J1644+57 could result
from the outer edge of a sub-parsec nuclear star cluster (e.g.,
\citealt{Carson+15}).  The high stellar densities of such a compact
star cluster could greatly enhance the TDE rate
(e.g.~\citealt{Stone&Metzger15}), possibly making this association
uncoincidental.

% We also note that \citet{Tchekhovskoy+2014} favor a smaller black hole
%($10^5-10^6 \Msun$) for J1644+57 based on the observed jet power, the jet
%shut-off time (if linked with the accretion becoming sub-Eddington), and an
%observed QPO in the x-ray light curve favor a smaller black hole {\bf
 % AG Gone mad!}.

@@

We have calculated steady-state models for the hot gaseous
circumnuclear media of quiescent galaxies, under the assumption that
gas is supplied exclusively by stellar wind mass loss and heated by
shocked stellar winds, supernovae and black hole feedback.  We compute
solutions for a range of different black hole masses, heating rates,
and observationally-motivated stellar density profiles.  Our numerical
results (Table \ref{table:models}) are used to verify and calibrate
analytic relationships (Appendix \ref{app:rs}, \ref{app:be}).  We use
the latter to explore systematically how the SMBH accretion rate
varies with black hole mass and the galaxy's SFH.  Our results for
$\dot{M}(\Mbh)$ are compared with observed trends of the nuclear X-ray
luminosities of quiescent SMBHs and low luminosity AGN.  They are also
used to address key questions such as the thermal stability of the
accretion flow.  Our conclusions are summarized as follows.

@@

  \item Including the effects of heat transport by electron conduction
    results in at most order unity changes to the key properties of
    the flow (e.g. stagnation radius, mass accretion rate, and thermal
    stability) for causal values of the conduction saturation
    parameter $\phi < 0.1$ (eq.~[\ref{eq:conduction}]).  Angular
    momentum is also unlikely to invalidate our spherically symmetric
    model on large scales, given the inferred stellar rotation rates
    near the stagnation radius ($\S\ref{sec:rotation}$). However, on
    small scales gas would encounter a centrifugal barrier which would
    cause much of the gas flowing inwards on large scales to be driven
    outwards in an outflow (as in \citealt{Li+2013}).

\item Radiative cooling has a more pronounced influence on the flow
  structure.  Cooling instabilities are likely to develop when the
  radiative cooling rate exceeds the gas heating rate near the
  stagnation radius.  This condition is approximately equivalent to
  the $t_{\rm cool} \gg t_{\rm ff}$ criterion advocated by
  \citet{McCourt+12} and \citet{Li&Bryan14a}.

@@

\item Stellar wind heating from the average SFH is sufficient to
  permit thermally-stable, steady-state accretion.  However, the
  resulting average Eddington-normalized accretion rate is predicted
  to increase weakly with $M_{\bullet}$ (Fig.~\ref{fig:miller}), in
  tension with the (also weak) downsizing trend of measured nuclear
  X-ray luminosities of early-type galaxies (e.g.,
  \citealt{Miller+15}).  However, these differences may not be
  discriminating given the huge uncertainty in translating the
  predicted gas inflow rate on parsec scales to the disk X-ray
  luminosity.

  Thermally stable accretion models can reproduce the observed scatter
  in nuclear X-ray luminosities at fixed $\Mbh$ of 2-4 orders of
  magnitude due to combination of (1) intrinsic scatter in the SMBH
  scaling relations; (2) stochastic variation in SFH from galaxy to
  galaxy, or from time to time due to burstiness in the SFH
  (e.g.~\ref{fig:NickPlot2}); (3) Variations in the amount of
  inflowing mass which is lost to small scale outflows from the SMBH
  accretion disk, due to variations in the angular momentum of the
  accreting gas (e.g.~\citealt{Li+13}).

\item The accretion rates of early-type galaxies are below the maximum
  value for thermally stable flow, $\dot{M}_{\rm TI}$
  (Fig.~\ref{fig:miller}), only for high X-ray radiative efficiencies.
  Therefore the current X-ray detections could instead be comprised
  mostly of nuclei undergoing outburst due to thermal instabilities.
  In this case there may exist a separate branch of low-$L_X$
  (undetected) thermally-stable nuclei, especially at low
  $M_{\bullet}$.  Such a bimodal population would call into question
  methods of extrapolating the $L_{\rm X}(M_{\bullet}$) relationship
  to low $M_{\bullet}$ in order to constrain the SMBH occupation
  fraction in low mass galaxies (\citealt{Miller+15}).

\item Low mass black holes grow in the low redshift universe over
  timescales comparable to the Hubble time (\citealt{Heckman+04}).  This
  accretion rates so required are too high to be consistent with the
  value predicted for the average SFH history and the maximum allowed
  for steady-state, thermally stable accretion
  (Fig.~\ref{fig:bh_growth}).  Low-$M_{\bullet}$ growth and AGN
  activity must instead be driven by a supply of gas external to the
  nucleus, e.g. due to galaxy mergers or thermal instabilities on
  larger, galactic scales (e.g.~\citealt{Voit+15}).

@@

We conclude by drawing attention to the following limitations of our
model.  First, by neglecting any large scale inflow of gas to the
nucleus and assuming the only source of gas feeding the black hole is
stellar wind mass loss from the local stellar population, our
estimates represent a lower bound on the time averaged gas density of
the CNM.  Our model therefore cannot account for AGN.

>>>>

\citet{BergerZauderer+:2012a} infer a flattening of the gas density
profile of the host of J1644+57 on radial scales $r \gtrsim 0.4$ pc
(their Fig.~6) which looks qualitatively similar to the shape density
profile we predict for core galaxies (Fig.~\ref{fig:profiles}, solid
line).  To obtain such a flattening on the inferred radial scales
would require a black hole of $\Mbh\sim 10^7 \Msun$.  However, this is
inconsistent with the break in the gas density seen at $\sim 0.6$ pc
by \citet{BergerZauderer+:2012a}, which should correspond to the break
radius of the {\it stellar density} profile, thus implying an
unrealistically small SMBH mass of $\Mbh\simeq 100\Msun$ according to
the mean $r_b-M_{\bullet}$ relationship \citep{LauerFaber+:2007a}.
Alternatively, the observed density break in J1644+57 could result
from the outer edge of a sub-parsec nuclear star cluster (e.g.,
\citealt{Carson+15}).  The high stellar densities of such a compact
star cluster could greatly enhance the TDE rate
(e.g.~\citealt{Stone&Metzger15}), possibly making this association
uncoincidental.  On the other hand, we note that some models for J1644+57  (e.g., \citealt{Tchekhovskoy+2014}) favor a much smaller black hole ($10^5-10^6 \Msun$) than we estimate would be required to produce the observed inner break.

% based on the observed jet power, the jet
%shut-off time (if linked with the accretion becoming sub-Eddington), and an
%observed QPO in the x-ray light curve favor a smaller black hole {\bf
 % AG Gone mad!}.

@@

We have calculated steady-state models for the hot gaseous
circumnuclear media of quiescent galaxies, under the assumption that
gas is supplied exclusively by stellar wind mass loss and heated by
shocked stellar winds, supernovae and black hole feedback.  We compute
solutions for a range of different black hole masses, heating rates,
and observationally-motivated stellar density profiles.  Our numerical
results (Table \ref{table:models}) are used to verify and calibrate
analytic relationships (Appendix \ref{app:rs}, \ref{app:be}).  We use
the latter to explore systematically how the SMBH accretion rate
varies with black hole mass and the galaxy's SFH.  Our results for
$\dot{M}(\Mbh)$ are compared with observed trends of the nuclear X-ray
luminosities of quiescent SMBHs and low luminosity AGN.  Our conclusions are summarized as follows.

@@

  \item Including the effects of heat transport by electron conduction
    results in at most order unity changes to the key properties of
    the flow (e.g. stagnation radius, mass accretion rate, and thermal
    stability) for causal values of the conduction saturation
    parameter $\phi < 0.1$ (eq.~[\ref{eq:conduction}]).  Angular
    momentum is also unlikely to invalidate our spherically symmetric
    model on large scales, given the inferred stellar rotation rates
    near the stagnation radius ($\S\ref{sec:rotation}$). 
%However, on
 %   small scales gas would encounter a centrifugal barrier which would
  %  cause much of the gas flowing inwards on large scales to be driven
   % outwards in an outflow (as in \citealt{Li+2013}).

\item Radiative cooling has a more pronounced influence on the flow
  structure.  Cooling becomes important when the
  radiative cooling rate exceeds the gas heating rate near the
  stagnation radius, where the gas is in nearly hydrostatic equilibrium (Fig.~\ref{fig:TI}).  This condition is approximately equivalent to
  the $t_{\rm cool} \gg t_{\rm ff}$ criterion for thermal instability advocated by
  \citet{McCourt+12} and \citet{Li&Bryan14a}.  The transition in the flow properties that occurs as heating is reduced may represent a true ``thermal instability" (hot ISM condensing into a cooler clouds), or it may simply represent an abruption transition from a steady inflow-outflow solution to a global cooling flow.  We leave distinguishing between these possibilities to future work.

@@

\item Stellar wind heating from the average SFH may be sufficient to
  permit thermally-stable, steady-state accretion, depending on the accretion efficiency of the SMBH.  However, the
  resulting average Eddington-normalized accretion rate is predicted
  to increase weakly with $M_{\bullet}$ (Fig.~\ref{fig:miller}), in
  tension with the (also weak) downsizing trend of measured nuclear
  X-ray luminosities of early-type galaxies (e.g.,
  \citealt{Miller+15}).  Thermally stable accretion models can reproduce the observed scatter
  in nuclear X-ray luminosities at fixed $\Mbh$ of 2-4 orders of
  magnitude due to combination of (1) intrinsic scatter in the SMBH
  scaling relations; (2) stochastic variation in SFH from galaxy to
  galaxy, or from time to time due to burstiness in the SFH
  (e.g.~Fig.~\ref{fig:NickPlot2}); (3) Variations in the amount of
  inflowing mass which is lost to small scale outflows from the SMBH
  accretion disk, due to variations in the angular momentum of the
  accreting gas (e.g.~\citealt{Li+13}).

\item However, for lower X-ray radiative efficiencies, the accretion rates of early-type galaxies are above the maximum value for thermally stable flow, $\dot{M}_{\rm TI}$
  (Fig.~\ref{fig:miller}).  This implies that current X-ray detections could instead be comprised
  mostly of nuclei undergoing outburst due to thermal instabilities.
  In such a case, there may exist a separate (usually undetected) branch of low-$L_X$ nuclei accreting stably.  The existence of such a bimodal population might call into question
  methods of extrapolating the $L_{\rm X}(M_{\bullet}$) relationship
  to low $M_{\bullet}$ to constrain the SMBH occupation
  fraction of low mass galaxies (\citealt{Miller+15}).

\item Low mass black holes grow in the low redshift universe over
  timescales comparable to the Hubble time (\citealt{Heckman+04}).  This
  accretion rates so required are too high to be consistent with the
  value predicted for the average SFH history and the maximum allowed
  for steady-state, thermally stable accretion
  (Fig.~\ref{fig:bh_growth}).  Perhaps unsurprisingly, low-$M_{\bullet}$ growth and AGN
  activity must instead be driven by a supply of gas external to the
  nucleus, e.g. due to galaxy mergers or thermal instabilities on
  larger, galactic scales (e.g.~\citealt{Voit+15}).

@@

We conclude by drawing attention to a few limitations of our
model.  First, we have neglected any large scale inflow of gas to the
nucleus by assuming the only source of gas feeding the black hole is
stellar wind mass loss from the local stellar population.  Our
estimates represent a lower bound on the time averaged gas density of
the CNM.  Our model cannot account for AGN.
